---
author: ""
---

# Suggested resources {.unnumbered}

## Git for beginners

### Pro Git

The Pro Git book is a standard resource for learning Git, offering a comprehensive and in-depth guide to version control with Git. 
Written by Scott Chacon and Ben Straub, the book covers everything from the basics of Git to advanced topics like branching strategies, workflows, and Git internals. 
The book is available for free online, making it an excellent resource for anyone looking to deepen their understanding of Git.
However, the Pro Git book can be dense, and some sections may feel overwhelming for newcomers. Additionally, while it’s very thorough, users looking for quick, specific answers might find the book's extensive coverage a bit too detailed. 
Despite these drawbacks, the Pro Git book is one of the best resources available for mastering Git.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| output: asis
bibtexkeys = c("chacon2014")
knitr::kable(ref_table(bibtexkeys), format = "markdown")
```

## GitHub 

### Official GitHub documentation

The official GitHub resources are essential for anyone working with Git and GitHub. 
They offer clear, detailed guidance on using GitHub for version control, collaboration, and project management, covering everything from basic Git commands to advanced features like Actions, Pull Requests, and Issues. 
The documentation is well-organized and user-friendly.
Regular updates ensure that the content stays relevant with the latest GitHub features and best practices.
A downside is that the documentation can sometimes be overwhelming due to its depth and the wide range of topics it covers.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| output: asis
bibtexkeys = c("GitHub2023")
knitr::kable(ref_table(bibtexkeys), format = "markdown")
```

