---
author: ""
---

# Style Guide {.unnumbered}

```{r}
#| eval: true
#| include: false
library("magrittr")
library("data.table")
```

We have compiled a set of guidelines to keep a consistent style across all chapters of the book.

## Summary

1. [Write one sentence per line](#write-one-sentence-per-line)
1. [Use consistent language](#use-consistent-language)

## Write one sentence per line

Please write each sentence on a new line.
Having each sentence on a new line will make no difference to how the text is displayed, there will still be paragraphs, but it will mean that any pull requests will be easier to check; the changes will be on a single line instead of somewhere in a paragraph.
Consider the example below.

```markdown
Today you are you, that is truer than true. There is no one alive who is youer than you. - Dr Seuss
```

A pull request on this correcting it to have a ‘.’ after Dr would show as a change to the whole paragraph.
Contrast this with the next example which will be displayed online in the exact same way, but would see a change to a single line.

```markdown
Today you are you, that is truer than true.
There is no one alive who is youer than you.
- Dr Seuss
```

## Use consistent language

We try to use consistent language: 

```{r}
#| eval: true
#| echo: false
variables <- yaml::read_yaml(here::here("_variables.yml"))
dt_input <- data.table::rbindlist(variables$language, fill = TRUE)
dt_output <- dt_input
```

```{r}
#| eval: true
#| echo: false
#| output: asis
dt_print <- dt_output %>%
  knitr::kable(., format = "markdown")
dt_print
```

## Acknowledgements & Attribution

- @community2022: [Style Guide](https://the-turing-way.netlify.app/community-handbook/style.html)
