---
author: ""
categories: [beginner, basics]
abstract: |
  In this chapter you will learn about the power of branches. Create, manage and merge branches like a pro!
---

# Branches

::: callout-note
### Overview

{{< fa clock >}} Duration: min
:::

::: {.column width="100%"}
{{< fa bullseye >}} Objectives

-   Knowing purpose and benefits of using branches in Git
-   Creating and switching between branches
-   Merging branches and resolving conflicts

:::

## Acknowledgements
```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| output: asis
bibtexkeys = c("GitHub2023", "chacon2014")
knitr::kable(ref_table(bibtexkeys), format = "markdown")
```

## Cheatsheet

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
knitr::kable(table_cheatsheet(name = "branches"), format = "markdown", row.names = FALSE)
```

## {{< fa dumbbell >}} Exercises 

-   Create a new Branch called feature/newveganreceipe 

-  Add a vegan recipe to your recipe text file

-   Merge it with your main branch and delete it afterwards

## {{< fa code-branch >}} Why branches?

Git branches are like separate workspaces within a Git repository. 
They let you work on different stuff without messing up the main code. 
You can visualize branches as shown in @fig-branches.

![Image from [Chapter "Git Branch"](https://www.atlassian.com/git/tutorials/using-branches) of the [Atlassian "Become a git guru" tutorials](https://www.atlassian.com/git/tutorials), used under a [Creative Commons Attribution 2.5 Australia License](https://creativecommons.org/licenses/by/2.5/au/) ([direct link](https://www.atlassian.com/git/tutorials/using-branches))](../static/branches.svg){#fig-branches}

Each branch keeps its own history, so you can develop things in parallel and easily merge changes between branches when you're ready. 
It's great for trying out new features, fixing bugs, or experimenting without causing chaos in the main development branch. 
Scientists might use branches to collaborate on research papers, exploring different parameters while running experiments or to test different data processing techniques or visualization approaches during data analysis.
Git makes it easy to switch between branches and handle conflicts when merging.
With branches, you can work independently and keep your code organized.

## Creating a new branch

By default your work is on the initial branch called **master** or **main**. 
We recommend changing your initial branch name to **main**, as discussed in the [installation](installation.qmd)  and [setup](setup.qmd) chapters.
You can check which branch you are on using `git branch`.

```{bash}
git branch 
```

should return the following output:

```{bash}
* main
```

To create a new branch, use `git branch` followed by the wished branch name.

```{bash}
git branch feature
```

Now you can use `git branch` again to confirm the creation of the new branch.

```{bash}
git branch 
```

should return the following output:

```{bash}
  feature
* main
```

::: {.callout-tip title="Common git branch flags" collapse="true"}
`-a` or `--all`: Lists both local and remote branches.

`-r or --remote`: Lists only remote branches.

`-d` or `--delete`: Deletes a specified branch. For example, git branch -d feature-branch.

`-D` or `--force`: Force deletes a branch, even if it has unmerged changes.

`-m` or `--move`: Renames a branch. For example, git branch -m old-branch new-branch.

`-c` or `--copy`: Creates a new branch by copying an existing branch. For example, git branch -c existing-branch new-branch

`-v` or `--verbose`: Shows more information when listing branches, including the last commit message.

`--contains`: Filters branches that contain a specific commit.
:::

## Switching branches

To switch to the newly created branch, you can use `git checkout` or `git switch`.

For example, enter the following command to switch to the newly created `feature` branch:

```{bash}
git switch feature
```

Should receive an output similar to this:

```{bash}
M	example.txt
Switched to branch 'feature'
```

Now you can normally work on this branch using the staging/committing workflow.

::: {.callout-tip title="What is the difference between `git checkout` and `git switch`?" collapse="true"}
Prior to Git version 2.23, the `git checkout` command was the primary way to switch branches. 
It allows you to move to a different branch and update your working directory to reflect the state of that branch. 
In Git 2.23 and later versions, the `git switch` command was introduced as a safer and more intuitive alternative to "git checkout". 
It simplifies the branch-switching process and provides clearer feedback in case of errors.
:::

## Merging branches

After you worked on your new branch you might want to incorporate your work into the `main` branch. 
You can do this using `git merge`command. 
Switch to your main branch (using `git switch` or `git checkout`) and then use:

```{bash}
git merge feature
```

If you have not yet done any changes you should get the following output:

```{bash}
Already up to date.
```

If you have changed the file(s) in your folder, you should get an output similar to:

```{bash}
Updating 555ba0c..994bb8d
Fast-forward
 
 example.txt    |   4 ++++

 1 file changed
```

## Merge conflicts

Merge conflicts occur when Git is unable to automatically reconcile differences between two branches during a merge or pull operation. 
This happens when both branches have made conflicting changes to the same part of a file. 
Git can't determine which version of the code should take precedence, and as a result, it requests manual intervention from the developer. 
It's like when two people try to edit the same part of a file at the same time, and Git gets confused about whose changes to keep. 
So, it raises its hand and asks you, the developer, to step in and help.

### Avoiding merge conflicts

Merge conflicts can be a bit of a headache. 
While it may not be possible to entirely avoid merge conflicts, employing good development practices can minimize their occurrence. 
Regularly pulling changes from the remote repository into your local branch (using git pull) and keeping your branch up-to-date can help identify and resolve conflicts early.

### Resolving merge conflicts

When a merge conflict arises, you should not view it as a limitation of Git but rather as a helpful feature. 
Git is like the referee saying, "Hey, programmers, sort out your differences!"

If you try to merge with conflicting changes in a file the output looks similar to this:

```{bash}
Auto-merging example.txt
CONFLICT (content): Merge conflict in example.txt
Automatic merge failed; fix conflicts and then commit the result.
```

Git will highlight the conflicting parts in the affected file(s)e, and you must choose the desired changes or modify them to create a consistent version. 
To do this open up the file(s) with conflicts in a text editor, and you'll see the conflicting sections marked with <<<<<<<, =======, and >>>>>>>.
Manually edit the conflicting sections to the desired state, removing the conflict markers. 
Decide which changes to keep and which to discard.

After editing the file, save it and use git add to stage the resolved changes:

```{bash}
git add example.txt
```

Once all conflicts are resolved, proceed with the commit e.g.:

```{bash}
git commit -m "Resolve merge conflict"
```

Your branches should now be merged and your conflict solved.

::: {.callout-tip title="Common `git merge` flags" collapse="true"}
`--abort`: Abort the current merge operation and reset the branch to its pre-merge state.

`--continue`: Continue the merge process after resolving conflicts.

`--log`: Include the log message from the commit(s) being merged in the new merge commit.
:::

### Types of merges

Git supports different types of merges to cater to various situations in software development. 
Each merge type has its purpose and advantages, allowing developers to choose the most suitable approach based on their specific project needs and workflow.

#### Standard merge

In a standard merge, Git takes each commit from the merged branch and adds them to the history of the base branch with their original time stamps. 
It creates a new "merge commit", a special record that marks when the merge happened, showing that the branches are now combined.

```{bash}
git merge 
```

#### Fast Forward Merge

If no new commits were made to the base branch since the feature was created, Git automatically does a "Fast Forward Merge."
It's like the standard merge, but no new merge commit is created. It's as if you made the commits directly on the base branch.
Since the base branch remained unchanged, there's no need to mark a separate merge in the history. You can ensure a fastfoward merge by using the `--ff-only`flag.

```{bash}
git merge --ff-only
```

If you do not want a fastfoward merge, you can ensure a merge commit with the `--no-ff` flag.

```{bash}
git merge --no-ff
```

#### Squash and Merge

When you squash a branch, Git combines all commits of the branch into a single commit.
This new commit is then added to the history, but the individual commits that were part of the branch are not retained or preserved separately.
Instead, the squashed commit represents the collective changes made in the branch, creating a cleaner and more concise history.

```{bash}
git merge --squash
```

::: {.callout-tip title="Common git rebase flags" collapse="true"}
`--interactive` or `-i`: Start an interactive rebase, allowing you to edit, reorder, or squash commits interactively.

`--continue`: Continue the rebase after resolving conflicts or editing commits during an interactive rebase.

`--abort`: Abort the current rebase operation and return the branch to its original state before the rebase.

`--skip`: Skip the current commit during an interactive rebase.

`-p` or `--preserve-merges`: Preserve merge commits during the rebase.

`--autosquash`: Automatically squash commits marked with "squash" or "fixup" in their commit message during an interactive rebase.
:::

## Deleting Branches

Once a branch has served its purpose and is no longer needed, it can be deleted.
Removing branches that are no longer active helps maintain a clean and manageable branch structure.
To delete a branch you can use the `git branch` command, followed by a `-d` flag.

```{bash}
git branch -d feature
```

should get an output similar to:

```{bash}
Deleted branch feature (was 3343f36).
```

You can only delete a branch, you are not currently on.

## Branches best practices

### Naming branches

Adopt a consistent naming convention for branches that accurately reflects their purpose.
This can include prefixes like `feature/`, `bug/`, or `hotfix/` followed by a descriptive name.
Clear and standardized branch names make it easier to identify the purpose of each branch.

### Branch Lifecycle

Create branches with a specific goal or task in mind, and aim to keep them short-lived.
Long-lived branches can lead to conflicts and make it harder to merge changes.
Once the task is complete or the feature is integrated, consider deleting the branch.

### Branch Protection

When using Git hosting platforms like GitHub or GitLab, it's a good idea to enable branch protection rules for your critical branch(es), typically your `main` or `master` branch.
By protecting it, you ensure that only reviewed and/or tested code gets merged into this branch, it also prevents accidental or unauthorized changes from being merged directly.

#### How to protect the Main Branch on GitHub

Go to your repository on GitHub and click on "Settings."
In the left sidebar, select "Branches." Under "Branch protection rules," click on "Add rule."
In the "Branch name pattern" field, enter the name of your main branch.
Enable the options you want to enforce for the main branch. Common options include:

**Require pull request reviews:** Require one or more approving reviews before changes can be merged.

Optionally, you can enable other protections, such as preventing force pushes or deleting the branch.

Click "Create" to save the branch protection rule.
