---
image: ../static/
categories: [intermediate]
abstract: |
  How does Git work under the hood?
engine: knitr
execute:
  eval: false
---

# Git internals
 
::: {.callout-tip appearance="minimal"}
<h5>Learning Objectives</h5>

:::

::: {.callout-note appearance="minimal"}
<h5>Exercises</h5>

:::

## Introduction

Purpose of the chapter

Concept of Git internals 


## `.git` folder

Significance, Location...

### `HEAD`

### `config` file

### `description` file

### `objects` folder

### `refs` folder

### `hooks` folder

### `logs` folder

## How Git stores data

### Hashes

### Blobs

### Commits

### Pack files

## Practical Tips

### Inspect Objects

### Repairing `.git` folder

### Exploring `.git` folder

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| output: asis
bibtexkeys = c("GitHub2023", "chacon2014", "community2022", "amin2023")
knitr::kable(ref_table(bibtexkeys), format = "markdown")
```

## Cheatsheet

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
knitr::kable(table_cheatsheet(name = "branches"), format = "markdown", row.names = FALSE)
