---
categories: [beginner, basics, setup]
abstract: |
  In this chapter, we will configure Git!
---

# Setup

::: {.callout-tip appearance="minimal"}
<h5>Learning Objectives</h5>
{{< include ../objectives/_objectives-setup.qmd >}}
:::

## Configuring Git

Once you [installed Git](installation.qmd), it is time to set it up.
We will use the the [command line](command-line.qmd) to set up and configure Git.
The `git config` command is used to get and set configuration. 
It allows you to customize Git according to your preferences and requirements. 
To look at all the different configuration commands you can use

```{bash}
git config
```

This chapter will focus on the most important and / or necessary configurations.

::: {.callout-tip title="What are the different configuration levels?" collapse="true"}
Git configuration values can be stored at **three different configuration levels**:

**System-level configuration**: This level applies configuration values to every user on the system and all their repositories.
To read or write configuration values at this level, you need administrative or superuser privileges.

**User-level configuration**: This level is specific to an individual user.
Configuration values set at this level will be applied to all repositories owned by the user.

**Repository-level configuration**: This level is specific to a particular Git repository.
The configuration values set at this level override values set at the system-level and user-level configurations for that specific repository.
By default, Git reads from and writes to this file when you use the `git config` command without any additional options.
:::

### {{< fa person >}} Identifying yourself

Before you start using Git, it's important to set your username and email address. 
This information is crucial because Git uses it to identify the author of each change you track with Git. 
By associating your changes with your username and email, it allows others (including future you) to see who made the changes which facilitates transparency and collaboration within a project.
You will be able connect Git to remote repositories like GitHub, using this email address. 
So it makes sense to set your email address to the one you use (or will use) to create a GitHub account.

Setting your username and email address is a one-time setup process, but you can always update them later, if needed. 
To perform this configuration, you can use the command line:

```{bash}
git config --global user.name "Jane Doe"
git config --global user.email jane@example.com
```

Make sure to replace `"Jane Doe"` with your own username and `jane@example.com` with your own email address.

::: {.callout-tip title="FAQ: Which name should I use for Git?" collapse="true"}
It can be your **GitHub username**, your **real name**, or something informative about you.
This name will be attached to the changes you track with Git, so **choose something that makes sense** to others and yourself in the future.
:::

### {{< fa code-branch >}} Configure the default name of the initial branch


Git uses a concept called "branches" to enable working on different versions of the same project. 
(More on this will be explained later in the book.)

On most installations of Git, by default, the name "master" for the initial branch is used when creating a new repository.
However, there has been a growing movement to transition to using "main" as the default branch name instead.
One of the main motivations behind this change is to promote inclusivity and remove offensive terminology from the default Git workflow (you can read more about the reasons [here](https://sfconservancy.org/news/2020/jun/23/gitbranchname/) or  [here](https://about.gitlab.com/blog/2021/03/10/new-git-default-branch-name/)). 
To configure the default branch name to "main" in Git, you can use the following command

```{bash}
git config --global init.defaultBranch main
```

## {{< fa gear >}} Check your settings

You can verify if your configurations were set correctly using the following command:

```{bash}
git config --list
```

This should create a long output list, in which you should find your set configurations, for example:

```{bash}
user.name=Jane Doe
user.email=jane@example.com
init.defaultbranch=main
```

## {{< fa circle-question >}} Calling for Help

The `--help` option in Git is a command that provides you with access to the built-in Git documentation and help resources.
When you run a Git command followed by `--help`, Git displays information about that command, including its usage, available options, and a brief description of its functionality.

For example, the following command will open up a browser in the command line containing information about the `git config` command.

```{bash}
git config --help
```

To quit the browser use `q`.

While the command provides extensive documentation and information about Git commands, it is arguably not the most beginner-friendly resource.

## {{< fa pen >}} Text editor

Sometimes, Git might open up a text editor inside the command line, and it can be annoying if you end up in an editor you don't like or don't know how to use. 
The standard text editor on macOS and when using "Git for Windows" is called [Vim](https://www.vim.org/), which is widely known for being difficult to use. 
You can find more information about the challenges of using Vim (and some funny memes) [here](https://stackoverflow.blog/2017/05/23/stack-overflow-helping-one-million-developers-exit-vim/).
To check out which text editor Git is using, you can use:

```{bash}
git config --get core.editor
```

If you have not changed your text editor yet, the output should be empty.
This means that Git is using the command line´s default editor, which should be **Vim** if you are using a Mac or Git for Windows. 
To make sure, you can use:

```{bash}
git var GIT_EDITOR
```

If the output is: `vim` or `vi`, your systems default editor is indeed Vim/Vi.
(Vim is an improved version of Vi, but there are virtually no differences between them for our usecase.)

In this situation, you can either pick up on the basics of Vim (see below) or simply change the text editor.
Also note that you will typically not have to use the text editor in the command line a lot.

### Vim Basics

To start [Vim](https://www.vim.org/), open the command line and type `vim` followed by the filename of the file that you want to edit (for example, `vim file.txt`).
In Vim, you have two main modes: "Insert" mode for typing and "Normal" mode for navigation and command execution.
Inside Vimn to switch from "Normal" mode to "Insert" mode, press `i` for inserting text.
To save your changes and exit Vim, press `Esc` to enter "Normal" mode, then type `:w` to save, and `:q` to quit.
Vim's strength lies in its powerful commands for text manipulation, such as copying, pasting, and searching, which are executed in "Normal" mode.

For a more detailed and interactive tutorial, you can check out [openvim.com](https://www.openvim.com).

### Changing your text editor

If you want to change the standard text editor, to one you are more comfortable with, there a lot of alternatives like [Nano](https://www.nano-editor.org/docs.php), [Visual Studio Code](https://code.visualstudio.com) or [Notepad++](https://notepad-plus-plus.org/downloads/).
If you have decided to use a different text editor you can use the following command (replace `editor-name` with the name of your preferred text editor):

```{bash}
git config --global core.editor editor-name
```

From now on, when Git requires you to interact with a text editor, it will use the one you have set.
If you ever want to change your Git editor in the future, you can repeat the above steps and specify the new editor command or path.

## Unsetting configuration

If you want to undo a configuration setting, you can use the `--unset` in the `git config` command.
For example, to unset the global configuration of your preferred text editor (`core.editor`) you can use the following command:

```{bash}
git config --global --unset core.editor 
```

## Acknowledgements and further reading

We would like to express our gratitude to the following resources, which have been essential in shaping this chapter.
We recommend these references for further reading:

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| output: asis
bibtexkeys = c("bryan2023", "capes2023", "chacon2014", "koziar2023")
knitr::kable(ref_table(bibtexkeys), format = "markdown")
```

## Cheatsheet

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
knitr::kable(table_cheatsheet(name = "config"), format = "markdown", row.names = FALSE)
```
